\section{Distributives}
Rather than describing nouns by their number, \langname\ describes nouns as either distributive or non-distributive.
Uses of the distributive form can be roughly divided into the following categories:

\begin{itemize}
  \item To form collective plurals
  \item To form plurals
  \item For prototypical distributive use
  \item To make generalizations
\end{itemize}

\subsection{Collective vs.\ collective plural}
Inherently dual nouns and collective nouns \nativeparen{物有二} may be modified by \DST\ to refer to multiple \detail{complete} sets.

\begin{examples}
  \baarucols{2}
  \ex
    \script 手
    \romanization heu
    \gloss
      ∅- & NDST \\
      heu & hand \\
    \tr (Both) hands
    \not (Two) hands
  \ex
    \script 各手
    \romanization meheu
    \gloss
      <me>heu & <DST>hand \\
    \tr Sets of hands
\end{examples}

\subsection{As a prototypical distributive}
\begin{example}\label{ex:dst-pred-cop}
  \script 各体鳥立垢
  \romanization menula sipu ta feisa
  \gloss
    <me>nula & <DST>body \\
    sipu & bird \\
    ta & ta \\
    feisa & filth \\
  \tr These bird carcasses (of many types) are so gross.
\end{example}

This sentence may appear to be a generalization of the qualities of bird carcasses.
However, generalized readings are more common with verbs inherently descriptive of opinion, rather than the predicational copular phrase \labelcref{ex:dst-pred-cop} presents.
\Cref{ex:dst-pred-cop} also acknowledges that under different circumstances, the bird carcasses \detail{may be perceived differently}.

\begin{example}\label{ex:ndst-pred-cop}
  \script 体鳥立垢
  \romanization nula sipu ta feisa
  \gloss
    nula & body \\
    sipu & bird \\
    ta & ta \\
    feisa & filth \\
  \tr These bird carcasses are so gross.
\end{example}

Being the non-distributive form of \labelcref{ex:dst-pred-cop}, \cref{ex:ndst-pred-cop} refers to a single type of carcass. Number is unspecified, instead referring to \detail{specific referents} established with prior context.

\subsection{Making generalizations/expressing opinions}
As mentioned previously, the use of \morphtext{me} to express opinion is mostly confined to its use alongside verbs describing emotion or state of mind.

\begin{examples}
  \ex
    \romanization a'au a'aeu kita'e auka
    \gloss
      a'\allo & ERG \\
      au & 1 \\
      a'aeu & enjoy\_presense \\
      ∅- & NDST \\
      kita & tree \\
      -'e & ASSOC \\
      auka & oak \\
    \tr I like those oak trees.
  \ex\label{ex:generalization-trees}
    \romanization a'au a'aeu mekita'e auka
    \gloss
      a'\allo & ERG \\
      au & 1 \\
      a'aeu & enjoy\_presense \\
      <me>kita & <DST>tree \\
      -'e & ASSOC \\
      auka & oak \\
    \tr I like oak trees.
\end{examples}

Generalized statements are often made by omitting personal pronouns, and optionally emphasizing the topic of the opinion.

In situations such as \labelcref{ex:generalization-trees} where the agent is making the generalization, the process of emphasizing the opinion is quite simple, because the patient can simply be fronted:

\begin{example}
  \romanization mekita'e auka a'aeu
  \gloss
    <me>kita & <DST>tree \\
    -'e & ASSOC \\
    auka & oak \\
    a'aeu & enjoy\_presense \\
  \tr Oak trees are pleasant.
\end{example}
\filbreak

This changes slightly when the agent is the opinion's topic. If further fronting cannot occur, \native{立} is used to emphatically mark agency:

\begin{example}
  \romanization ta simesikama aseisi
  \gloss
    ta & ta \\
    si\allo & ERG\\
    <me>sikama & <DST>buzzing\_insect \\
    asei & scare \\
    -si & INV \\
  \tr Insects are scary.
  \lit Insects scare (me).
\end{example}

An exception to the verbal restrictions surrounding generalization occurs when expressing opinions about the \detail{attributes} of an object.
\aside{Qualities which can be described this way are idealized or culturally significant.
Opinions on these qualities are seen as more important, and the choice to describe them this way gives the opinion more weight than otherwise imparted by a more peripherastic construction.}
One caveat is that the attribute must be one which can be described with the associative particle, \rom{'e}.

\begin{example}
  \script 各木亜脊立煩
  \romanization mekita'e pala ta sika
  \gloss
    <me>kita & <DST>tree \\
    -'e & ASSOC \\
    pala & height \\
    ta & ta \\
    sika & worry \\
  \tr Tall trees are foreboding.
\end{example}
\subsection{Singular vs.\ plural}
The plural reading of \morphtext{me} occurs with personal pronouns, foreign loans that are not \native{物有二}, or loans which have not been sufficiently nativized, as with \labelcref{ex:plural-pronouns}.
Established loans, such as \rom{i'isi} \trans{foreigner} follow the prototypical distibutive reading.

\begin{columns}[cols.markup=\mutations]\label{ex:plural-pronouns}
  \cols leunu & meleunu & cactus/cactuses & (Lérvima \rom{lonón}) \\
  \cols neke & meneke & goat/goats & (Epe \rom{ŋehke})
\end{columns}
