\chapter{Additional Example Sentences}
\ex
\begingl
\glpreamble mapiʔe ai pa tasaʔe kiɸe tamuka asuhi
\pronounced{ˈma.pi.ʔə ˈai ˈpa ˈta.sa.ʔə ˈki.ɸə ˈta.mu.ka ˈa.su.hi}\endpreamble
\nogloss{\lbrack}
mapi[whole]
ʔe[\textsc{assoc}]
ai[one]
\nogloss{\rbrack}
pa[now]
tasa[chaos]
ʔe[\textsc{assoc}]
kiɸe[place]
tamuka[\textsc{cert}]
asuhi[increase]
\glft `In isolation, chaos surely follows.'\footnotemark\\
\lit{The whole is [made from] one, so chaos must increase.}\smoyd{1}{https://www.reddit.com/r/conlangs/comments/28vg3k/i_just_used_up_5_minutes_of_your_day_day_1/}
\endgl
\xe

\footnotetext{The original text was: `In an isolated system, entropy can only increase.'}

The translation of `isolated system' is not terribly transparent.
I've taken some liberties with the connotations of the phrase while translating, such that this phrase is less about literal entropy and more about the fragmentation of a group.
Countable nouns are associated with a number in the same fashion adjectives are.
Thus, \langword{mapiʔe ai} refers to a group of only one member, and therefore one which exists in isolation.

The coordinating conjunction \langword{pa} considers the second clause a direct result of the first.
This contrasts with \langword{alete,} which considers the event to have happened due to a culmination of the first among other factors.
In other words, while \langword{alete} creates a much more definitive sense of causation, \langword{pa} functions more aptly when it comes to sequencing.

Because I have reinterpreted this phrase as referring to a group, I have also described the chaos in a particular way.
Namely, considering that the members of the group are assumed to still be alive, it is the place which they meet which is considered to be chaotic.
Alternatively, the people themselves could have been described as having chaos grow within them, but this would have the connotation that the people themselves were driving the group apart rather than some external factor.

\langword{Tamuka} is used for a few reasons.
These sorts of evidentials in some contexts, can play the role of what would be rendered with modal verbs in English.
In this particular instance, \langword{tamuka} also helps cement the idea that this course of action is the only possible one (ie. all others would be therefore be completely unfathomable).

\ex
\begingl
\glpreamble menexesi meauna lu mapi muhipese
\pronounced{ˈme.nə.xə.si ˈme.au.na ˈlu ˈma.pi ˈmu.hi.pə.sə}\endpreamble
\nogloss{\lbrack}
menexesi[\textsc{pl:}star]
meauna[\textsc{pl:}moon]
lu[\textsc{obv}]
mapi[whole]
\nogloss{\rbrack}
muhipese[\textsc{pst.pfv:}tame]
\glft `The stars, the moon, they have all been blown out.'\\
\lit{The stars, the moon, they were all tamed.}\smoyd{2}{https://www.reddit.com/r/conlangs/comments/28y706/i_just_used_up_5_minutes_of_your_day_day_2/}
\endgl
\xe

The first thing to note in this translation, is that when noun phrases are not joined by \langword{ke}, and instead listed as is done here, each noun is inflected separately.
Another thing to note is that the world which \langname\ is spoken in has two moons, therefore a plural form makes more sense rather than the original singular.

The particle \langword{lu} is used so that both referents can be considered a single unit, and therefore both are considered in reference to \langword{mapi.}
Unlike other things, \langword{mapi} does not require an \langword{ʔe} to associate \langword{mapi} with any particular `whole'.
This is because there is not one particular whole being described, but rather a reference to an already existing complete group of celestial objects.

Because the stars and the moons were introduced as a patient of some verb, we may be expecting the agent to follow the verb, as it normally would in a situation with a focused patient argument.
In this particular instance however, no agent is present, and no \langword{ta} Is used to introduce any sort of reflexive reading.
\langword{Hipese} is a labile verb, meaning the sole argument is rightfully a patient (in agreement previously established case marking).
As a convenient consequence of \langword{hipese} being labile, the translated text renders much more precisely to the English.

The choice of verbs itself is notable, as this is used often in reference to exerting control over a natural forces.

\ex
\begingl
\glpreamble ɸene suɸai uʔtwaw ɸaipa ʔutawa tu akatewiʔe
\pronounced{ˈɸe.nə ˈsu.ɸai ˈuʔ.ta.wa ˈɸai.pa ˈʔu.ta.wa ˈtu ˈa.ka.te.wi.ʔə}\endpreamble
ɸene[live]
su-ɸai[without/\textsc{aug-neg}]
uʔtawa[die]
ɸaipa[unless/\textsc{neg-}pa\footnotemark]
ʔutawa[die]
tu[\textsc{2sg}]
akatewi-ʔe[warrior-ʔe]
\glft `Life is pointless... unless you die a warrior.'\\
\lit{Life will be without death, unless you die [like] a warrior.}\smoyd{3}{https://www.reddit.com/r/conlangs/comments/291tml/i_just_used_up_5_minutes_of_your_day_day_3/}
\endgl
\xe

\footnotetext{See X for a full explanation of \langword{pa.}}

\ex
\begingl
\glpreamble
\pronounced{ˈ}\endpreamble
\glft `These leviathan like creatures glimmer in the rays of the blast as they swarm and beat at the wreckage'\\
\lit{These leviathan like creatures glimmer in the rays of the blast as they swarm and beat at the wreckage.}\smoyd{4.2}{https://www.reddit.com/r/conlangs/comments/294syw/i_just_used_520_minutes_of_your_day_day_4/}
\endgl
\xe

\ex
\begingl
\glpreamble
\pronounced{ˈ}\endpreamble
\glft `Outside the window is misty and dull'\\
\lit{Outside the window is misty and dull.}\smoyd{4.2}{https://www.reddit.com/r/conlangs/comments/294syw/i_just_used_520_minutes_of_your_day_day_4/}
\endgl
\xe