\chapter{Nominal}\label{ch:morpho-nom}
\section{Inflectional Morphology}\label{sec:morpho-nom-inf}
Inflectional morphology is very limited for nouns. The most marked nouns only inflect for case marking and plurality. However, certain constructions call for plural marking to be absent altogether, further reducing the variance in word shape.

\subsection{Case Marking}\label{sec:case-marking}
\langname 's case system is incredibly small. It consists solely of the cases used for morphosyntactic purposes --- the ergative and absolutive. The ergative case is marked with initial syllable reduplication, while the absolutive case remains unmarked. This becomes slightly less transparent when the reduplication triggers allophony and interacts with phonotactics.

\aside{\fleuron\ While phonological form is not the main focus of this section, the variation that can occur makes it useful. It \textit{is} true that much of the prose simply reiterates what is found in the pronunciation information found in the gloss; however, I consider this approach more useful for grasping the reasoning and the triggers behind these phonological changes.}

With \notabletext{CV-initial forms,} this is quite straightforward:

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble pane
    \pronunciation ˈpa.nə
    \gloss
      pane & person[ABS] \\
  \ex
    \preamble papane
    \pronunciation pa.ˈpa.nə
    \gloss
      pa\allo pane & ERG\allo person \\
\end{subexamples}

\notabletext{V-initial forms} are slightly different, as adjacent vowels cannot be identical. An epenthetic \phomtext{ʔ} is inserted to avoid this.

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble ana
    \pronunciation ˈa.na
    \gloss
      ana & eye[ABS] \\
  \ex
    \preamble a'ana
    \pronunciation aʔ.ˈa.na
    \gloss
      aʔ\allo ana & ERG\allo eye \\
\end{subexamples}

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble ele
    \pronunciation ə.ˈle
    \gloss
      ele & heaven[ABS] \\
  \ex
    \preamble e'ele
    \pronunciation əʔ.ə.ˈle
    \gloss
      eʔ\allo ele & ERG\allo heaven \\
\end{subexamples}

\phomtext{ə}-initial forms (such as the one above) are particularly interesting in that they most clearly show an interaction between reduplication and allophony. This form of reduplication can be analyzed as the morpheme \morphtext{V,}. As a result, \phomtext{əʔ.ə.lə} is phonetically \phontext{əʔ.ə.ˈle} rather than \phontext{əʔ.ˈe.lə}.

\notabletext{CVV-initial forms} undergo a slightly different process. The reduplication in this situation does not even apply to the entire syllable, but rather, only to the CV segment.

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble keuha
    \pronunciation ˈko.ha
    \gloss
      keuha & leaf[ABS] \\
  \ex
    \preamble kekeuha
    \pronunciation kə.ˈko.ha
    \gloss
      ke\allo keuha & ERG\allo leaf \\
\end{subexamples}

Consequently for LSA, the reduplicated vowel may be different than its phonetic realization in the stem due to an underlying diphthong.

\notabletext{VV-initial forms} work similarly to CVV forms, with the addition of an epenthetic \phomtext{ʔ} at the morpheme boundary:

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble auna
    \pronunciation ˈau.na
    \gloss
      auna & moon[ABS] \\
  \ex
    \preamble a'auna
    \pronunciation aʔ.ˈo.na
    \gloss
      aʔ\allo auna & ERG\allo moon \\
\end{subexamples}

As illustrated by the previous set of examples, this also triggers mono\-phthongization of the stem's initial syllable.

\subsection{Pluralization}\label{sec:pluralization}
Pluralization is very regularly marked through the infixation of \morphtext{me} between the reduplicated syllable of an ergative form and the stem. In absolutive forms, this gives \morphtext{me} the appearance of a prefix.

Let's revisit some of the previous examples with plural forms:

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble mepane
    \pronunciation mə.ˈpa.nə
    \gloss
      <me>pane & <PL>person[ABS] \\
  \ex
    \preamble pamepane
    \pronunciation pa.ˈme.pa.nə
    \gloss
      pa\allo <me>pane & ERG\allo <PL>person \\
\end{subexamples}

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble meana
    \pronunciation mə.ˈa.na
    \gloss
      <me>ana & <PL>eye[ABS] \\
  \ex
    \preamble ameana
    \pronunciation a.ˈme.a.na
    \gloss
      a\allo <me>ana & ERG\allo <PL>eye \\
\end{subexamples}

\begin{subexamples}
  \label{ex:double-schwa}
  \baarucols{2}
  \ex
  \preamble meele
  \pronunciation ˈme.lə
  \gloss
    <me>ele & <PL>heaven[ABS] \\
  \ex
    \label{ex:double-schwa-erg}
    \preamble emeele
    \pronunciation ə.ˈme.lə
    \gloss
      e\allo <me>ele & ERG\allo <PL>heaven \\
\end{subexamples}

\aside{Loanwords from other languages are commonly respelled with \orthotext{ee} for \phontext{e}. Older generations generally dislike this emerging tendency, as it reinforces foreign loans.}

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble mekeuha
    \pronunciation mə.ˈko.ha
    \gloss
      <me>keuha & <PL>leaf[ABS] \\
  \ex
    \preamble kemekeuha
    \pronunciation kə.ˈme.ko.ha
    \gloss
      ke\allo ke<me>uha & ERG\allo <PL>leaf \\
\end{subexamples}

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble meauna
    \pronunciation mə.ˈau.na
    \gloss
      <me>auna & <PL>moon[ABS] \\
  \ex
    \preamble ameauna
    \pronunciation a.ˈme.o.na
    \gloss
      a\allo <me>auna & ERG\allo <PL>moon \\
    \end{subexamples}

\baaruref{ex:double-schwa} is notable for being one of the few situations utilizing \orthotext{ee}. It is preserved in some words to distinguish otherwise opaque stress or as an etymological consequence such as in \baaruref{ex:double-schwa-erg}.

\section{Derivational Morphology}
\begin{description}
  \item[-la] derives from an adjective a noun meaning \transtext{composed from X}
  \item[-ku] constructs a verb with the meaning \nativetext{ewi-siha hasi X ta} \transtext{to happen as if by X, to appear as X}, where X is a noun
\end{description}

\subsection{-la}

The primary and perhaps most straightforward use of \nativetext{-la} is that of material composition:

\begin{examples}
  \baarucols{2}
  \ex
    \preamble lelu suxela
    \pronunciation lə.ˈlu ˈsu.xə.la
    \gloss
      lelu & spoon \\
      suxe & \MC2 processed\_wood-ADJZ \\
      -la & \\
    \tr wooden spoon
  \ex
    \preamble kele wasila
    \pronunciation kə.ˈle ˈwa.si.la
    \gloss
      kele & bowl \\
      wasi & river\_stone-ADJZ \MC2 \\
      -la & \\
    \tr stone bowl
\end{examples}

\nativetext{-la} should not be confused with the uses of the English preposition \transtext{of}.

\begin{example}
  \preamble ti'ika kitala
    \pronunciation ˈti.ʔi.ka ˈki.ta.la
    \gloss
      ti'ika & branch \\
      kita & raw\_wood-ADJZ \MC2 \\
      -la & \\
    \tr wooden branch
    \intended tree branch
\end{example}

This particular phrase calls for the associative particle \nativetext{'e}, which a\-chieves the desired semantic relationship. This will be further detailed in \ref{ch:morpho-adj}; the correct formation is illustrated here:

\begin{example}
  \preamble ti'ika'e kita
    \pronunciation ˈti.ʔi.ka.ʔə ˈki.ta
    \gloss
      ti'ika & branch \\
      -'e & ASSOC \\
      kita & tree \\
    \tr tree branch
\end{example}

Apart from this, \nativetext{-la} can also designate the origin of a person or idea. Notably, it refers to origination which is generic rather than circumstantial.

This distinction manifests in \langname 's strategies for expressing residency.

\aside{While it is expected that a \nativetext{la}-toponym refer to one's birthplace, this is not univerally true in colloquial usage.}
\aside{In the case of young children who grow up in a different place than they were born, \nativetext{la}-toponyms may shift to the more formative, newer location.}

%? Uakci what should I call 'la-forms' instead? Currently reads a bit weird
%?? TODO – ponder on this more ig

\begin{examples}
  \baarucols{2}
  \ex
    \preamble me ɸene wu ehaixa
    \pronunciation ˈme ɸə.ˈne ˈwu ə.ˈhai.xa
    \gloss
      me & 1sg \\
      ɸene & live \\
      wu & on\_top \\
      Ehaixa & NAME \\
    \tr I live in Ehaixa.
  \ex
    \preamble me ehaixala
    \pronunciation ˈme ə.ˈhai.xa.la
    \gloss
      me & 1sg \\
      Ehaixa & NAME-ADJZ \MC2 \\
      -la & \\
    \tr I am from Ehaixa.
\end{examples}

As mentioned above, \nativetext{-la} can also be used to describe the origin of an idea or concept. This can be used formally, in the sense of invention, or simply to describe where a thought comes from.

This is found primarily within conversational, reportative discourse, rather than in formal accounts or official notices.

\aside{Auku is most often translated as \transtext{hear}, though it is also used as a general sensory verb more akin to \transtext{feel}. However, \nativetext{auku} is not used in the context of emotion.}
\begin{subexamples}
  \ex
    \preamble pane auku tu (kini)
    \pronunciation ˈpa.nə ˈau.ku ˈtu (ˈki.ni)
    \gloss
      pane & person \\
      auku & hear \\
      tu & 2SG \\
      ( kini ) & this \\
    \tr Where did you hear that?
    \alt Who told you that?
  \ex
    \preamble nalexu tala
    \pronunciation ˈna.lə.xu ˈta.la
    \gloss
      Nalexu & NAME \\
      ta & PROX-la \MC2 \\
      -la & \\
    \tr Nalexu did!
\end{subexamples}

When \nativetext{-la} would be attached to names, particularly those of people, it is instead attached to the corresponding focus marker.

\aside{This stems from a widespread avoidance of inflecting proper nouns. Similar avoidance strategies can be found across a variety of constructions.}

\notabletext{Formal use of \nativetext{-la}} is typically reserved for description of relationships which are for the most part, indisputable. It can express a particular type of ownership most easily correlated with geographical location, but it is also used in a more political sense in reference to whole nations.

\begin{example}
  \preamble nauwa ateenala
  \pronunciation ˈnau.wa a.ˈte.na.la
  \gloss
    nauwa & flowing\_water \\
    Ateena & NAME-la \MC2 \\
    -la & \\
  \tr Ateenalen water
\end{example}

\subsection{-ku}

\aside{\nativetext{-ku} comes from a colloquial shortening of the verb \nativetext{ku'ime} \transtext{to attempt}.}

\nativetext{-ku} in isolation derives an intransitive verb, most often used to describe the manner by which domething occurred.

It's meaning has shifted since the construction originally diverged, resulting in the additional hedging found in the present definition. Rather than the outcome being unspecified, as with \nativetext{ku'ime}, \nativetext{-ku} generally imparts that the attempt did not go as expected. If successful, it will convey dishonesty or trickery in the resulting achievement.

\aside{\nativetext{Auku} is generally used as a sensory verb, but consequently it can be zero-derived to a noun roughly meaning \transtext{knowledge, experience, skill}.}

\begin{example}
  \preamble anaku, eusaku, alete auku'u
  \gloss
    ana & eye-VBZ \MC2 \\
    -ku & \\
    eusa & ear-VBZ \MC2 \\
    -ku & \\
    alete & thus \\
    auku & knowledge-VBZ \MC2 \\
    -'u & \\
  \tr I looked, I listened, but I couldn't find anything.
\end{example}

In this example, \nativetext{-ku} Simply conveys unsuccessful action. Notably, no overt negation is present. This is a feature of this derivation, especially in conjunction with \nativetext{alete}.

\begin{subexamples}
  \baarucols{2}
  \ex
    \preamble wa meme suxi!
    \pronunciation ˈwa mə.ˈme ˈsu.xi
    \gloss
      wa & AFF \\
      me\allo me & ERG\allo 1SG \\
      suxi & solve \\
    \tr I figured it out!
  \ex
    \preamble ta auku'u
    \pronunciation ˈta ˈau.ku.ʔu
    \gloss
      ta & PROX \\
      auku & knowledge-VBZ \MC2 \\
      -'u & \\
    \tr Of \textit{course} you did...
\end{subexamples}

This exchange highlights the negative connotations of \nativetext{-ku}, particularly in its use in conveying trickery. The second speaker believes the first is not smart enough for such critical thinking. While it \textit{is} a fairly generic insult, it is partially contextually sensitive.

%* Remember to note recent past use with alete + present tense

This derivation is most productively used within compound verbs.

\aside{\nativetext{Lasaka'ane} is derived from compounding of \nativetext{lasaka} \transtext{choice} + \nativetext{pane} \transtext{person}}

\begin{examples}
  \baarucols{2}
  \ex
    \preamble lasaka'aneku suxi
    \gloss
      lasaka'ane & council-VBZ \MC2 \\
      -ku & \\
      suxi & solve \\
    \tr to come to a mutual understanding
  \ex
    \preamble lasaka'aneku
    \gloss
      lasaka'ane & council-VBZ \MC2 \\
      -ku & \\
    \tr to debate
\end{examples}